Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-07-08 manuscript/manuscript.Rmd
updates ms with new abstract and adds crowdXsize interactions in equations

Abstract
--------
The ability of population models to skillfully forecast future states under climate change is constrained by the limited spatial and temporal extent of demographic data. This data is often limited because it is difficult and costly to collect. An alternative is to rely on aggregate, population-level data that is easier and less costly to collect. Doing so requires assuming that population-level data accurately represents the aggregate response of the individuals that actually respond to weather. We tested this assumption using population models of four Montana grassland species fit using individual and aggregated forms of the same data. We fit population models with interannual variation in vital rates explained, in part, by climate covariates and then perturbed observed climate to compare model forecasts. Population models based on individual-level demographic data outperformed the models based on aggregate-level data in terms of accuracy and precision. The two model types produced inconsistent forecasts when we perturbed climate. Thus, it seems that, at least for the species in this location, demographic data is necessary to pick up key climate drivers of vital rates like survival that are not well resolved in aggregated data. On a pessimistic note, our work shows that even with detailed demographic data, model forecasts are extremely uncertain. It is becoming clear the long time series (over two decades) are necessary to produce meaningful forecasts.

@@

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual level data we fit models for survival, growth, and recruitment of new individuals for each species. At the quadrat level we fit a single regression model for population growth. We describe the statistical models separately since fitting the models required different approaches. All models contain five climate covariate that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year. We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.

@@

\begin{align}
\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
\end{align}

where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species. $\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat group $Q$.

@@

\begin{align}
x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
\end{align}

@@

where $x_{jq,t}$ is the log of species' $j$ proportional cover in quadrat $q$ at time $t$ and all other parameters are as in the individual-level growth model (Eq. 3). The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.

@@

Conclusions
-----------
This work is not a critique of density-structured population models. In some cases and for certain species, population models based on aggregated data may prove useful and unbiased. However, our work here is the first comparison of population models based on individual and aggregated data in a climate change context. Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical to predicting species' responses to climate change. Thus, forecasts from aggregate level models should be viewed with caution and should never be unaccompianed by uncertainty.

>>>>

Abstract
--------
Forecasting future states of populations has taken on new urgency as the rate of climate change increases. Traditional plant population models have limited utility in this regard because they are based on detailed demographic data from small, localized plots. These models are difficult to scale up to spatial scales relevant to land managers that require such forecasts to make decisions. To overcome the data limitations of traditional population models, some have proposed population models based on population level, rather than individual level, data that is much easier to collect over broad spatial scales. Using such models violates a central assumption of ecology: individuals respond to weather, not populations to climate. Here, we test whether this assumption is important when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland. We parameterized two population models, one based on inidividual level data with three vital rates and one on an aggregated version of the same data (percent cover), with climate covariates and compared their accuracy, precision, and sensitivity to climate. The individual level model was more accurate and precise than the aggregate level model when predicting out of sample observations. The aggregate level model made countervailing forecasts to 1% climate changes when compared to the individual level model, indicating the aggregate level model produces incorrect forecasts. When comparing climate effects from both models, the aggregate level model tends to "miss" important climate effects from at least one vital rate for each species. It appears there is no short cut to forecasting climate change impacts on plant populations — detailed demographic data is essential. But, forecasts were very uncertain, so we advocate for a focus on new methods to collect demographic data more efficiently across environmental gradients in space and time.

@@

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual level data we fit models for survival, growth, and recruitment of new individuals for each species. At the quadrat level we fit a single regression model for population growth. We describe the statistical models separately since fitting the models required different approaches. All models contain five climate covariate that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year. We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions. Climate $\times$ size interactions are for climate main effects only, that is we do not include interactions among size and interacting climate effects.

@@

\begin{align}
\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
\end{align}

where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). Note that the vector of climate covariates (**C**) includes climate variable interactions and climate$\times$size interactions. We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species. $\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat group $Q$. We include a size$\times$crowding interaction effect ($\nu^{S}$).

@@

\begin{align}
x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
\end{align}

@@

where $x_{jq,t}$ is the log of species' $j$ proportional cover in quadrat $q$ at time $t$ and all other parameters are as in the individual-level growth model (Eq. 3). Again, note that the climate covariate vector (**C**) includes the climate$\times$cover interaction. The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.

@@

Conclusions
-----------
This work is not a critique of density-structured population models. In some cases and for certain species, population models based on aggregated data may prove useful and unbiased. However, our work here is the first comparison, to our knowledge, of population models based on individual and aggregated data in a climate change context. Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical to predicting species' responses to climate change. Thus, forecasts from aggregate level models should be viewed with caution and should never be unaccompianed by uncertainty. Given the importance of demographic data and its current difficulty to collect, we advocate for research on modern methods to collect demographic data more efficiently across environmental gradients in space and time. For example, recent advances in object recognition algorithms [@Liu2014] provide new opportunities to speed up data collection and digital processing.

### 2015-07-08 manuscript/manuscript.tex
updates ms with new abstract and adds crowdXsize interactions in equations

@@

The ability of population models to skillfully forecast future states
under climate change is constrained by the limited spatial and temporal
extent of demographic data. This data is often limited because it is
difficult and costly to collect. An alternative is to rely on aggregate,
population-level data that is easier and less costly to collect. Doing
so requires assuming that population-level data accurately represents
the aggregate response of the individuals that actually respond to
weather. We tested this assumption using population models of four
Montana grassland species fit using individual and aggregated forms of
the same data. We fit population models with interannual variation in
vital rates explained, in part, by climate covariates and then perturbed
observed climate to compare model forecasts. Population models based on
individual-level demographic data outperformed the models based on
aggregate-level data in terms of accuracy and precision. The two model
types produced inconsistent forecasts when we perturbed climate. Thus,
it seems that, at least for the species in this location, demographic
data is necessary to pick up key climate drivers of vital rates like
survival that are not well resolved in aggregated data. On a pessimistic
note, our work shows that even with detailed demographic data, model
forecasts are extremely uncertain. It is becoming clear the long time
series (over two decades) are necessary to produce meaningful forecasts.

>>>>

\newpage{}

@@

Forecasting future states of populations has taken on new urgency as the
rate of climate change increases. Traditional plant population models
have limited utility in this regard because they are based on detailed
demographic data from small, localized plots. These models are difficult
to scale up to spatial scales relevant to land managers that require
such forecasts to make decisions. To overcome the data limitations of
traditional population models, some have proposed population models
based on population level, rather than individual level, data that is
much easier to collect over broad spatial scales. Using such models
violates a central assumption of ecology: individuals respond to
weather, not populations to climate. Here, we test whether this
assumption is important when forecasting climate change impacts on four
perennial grass species in a semi-arid Montana grassland. We
parameterized two population models, one based on inidividual level data
with three vital rates and one on an aggregated version of the same data
(percent cover), with climate covariates and compared their accuracy,
precision, and sensitivity to climate. The individual level model was
more accurate and precise than the aggregate level model when predicting
out of sample observations. The aggregate level model made
countervailing forecasts to 1\% climate changes when compared to the
individual level model, indicating the aggregate level model produces
incorrect forecasts. When comparing climate effects from both models,
the aggregate level model tends to ``miss'' important climate effects
from at least one vital rate for each species. It appears there is no
short cut to forecasting climate change impacts on plant populations ---
detailed demographic data is essential. But, forecasts were very
uncertain, so we advocate for a focus on new methods to collect
demographic data more efficiently across environmental gradients in
space and time.
